\onehalfspacing
\phantomsection % this is to manually fix anchor
\addcontentsline{toc}{section}{\textit{Abstract}}
{\fontsize{16pt}{30pt}\selectfont \noindent\textbf{Abstract}}
\vspace{0.5cm}
{\fontsize{12pt}{18pt}\selectfont
% replace the paragraph below with your abstract

AI agents increasingly rely on external tools to extend their capabilities, yet existing systems lack efficient and secure OS-level support for managing tool invocation. 
Unrestricted access to Large Language Models(LLMs) and external tools introduces risks of resource misuse, security vulnerabilities, and unreliable execution, highlighting the need for AI-native OS support.

To address this, we present \texttt{linux-mcp}, a kernel-assisted control plane for tool invocation on Linux operating system.
It introduces a novel operating system architecture for serving AI agents by keeping semantic parsing and tool execution in user space while moving admission control into a Linux kernel module.
This design enables strong isolation via a unified daemon that mediates all external tool invocations, enforcing a strict OS-tool security boundary with minimal overhead.
Furthermore, the Model Context Protocol(MCP) abstraction enables seamless tool integration: external tools can be exposed to AI agents by simply specifying their functionality and parameters in a manifest, without additional implementation.

We evaluate \texttt{linux-mcp} against both a pure user-space baseline and a hardened user-space design across latency, throughput, attack resistance, and failure scenarios.
Our category-level evaluation confirms that the kernel-based design correctly enforces the specified control-plane invariants on every tested attack case (0 of 110 attempts succeed against the kernel path, versus 90.9\% and 36.4\% bypass rates against the two instrumented user-space baselines), and a boundary-condition supplement confirms that these invariants also hold against minimum-distance value forgeries and a concurrent-consume race (0 of 240 additional attempts succeed). The latency cost of kernel placement is small enough to be within the noise envelope of a VMware guest at small and medium payloads and measurably favourable to the kernel path at megabyte-class payloads, and throughput remains within the same practical range as the baselines under the tested multi-agent workloads.
The prototype exposes 43 tools across 14 applications and is publicly available at \url{https://github.com/xiheng724/linux_mcp}.

}
\vspace{0.5cm}
\addcontentsline{toc}{section}{\textit{Keywords}}
{\fontsize{16pt}{30pt}\selectfont \noindent\textbf{Keywords}}
\vspace{0.5cm}
{\fontsize{12pt}{18pt}\selectfont
% replace the paragraph below with your keywords

AI Agents, Operating Systems, Tool Invocation, System Security

}


